$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-08-01 Referee_Response.txt

Responses to First Referee:

We thank the First Referee for his/her comments. Here we directly respond to the following particular comments (with the referee's comment in quotes):

"Definition of Q_v and c_P It is quite difficult to see what the 
authors mean by $Q_v$ and $c_P$ from the text. I guess the definition 
of $n_\ell$ in page 2 should read $n_\ell = (1+\sigma^z_\ell)/2$ if 
$\sigma^z_\ell$ takes value $\pm 1$."

Yes this is correct--we appreciate the referee for catching this error.

@@

"The definition of $c_P$ is 
cryptic because "the even (odd) sector of the constraint" seems not 
defined. I was only able to see the assignment of $Q_v$ to each 
configuration from TABLE I. It would be better to touch on TABLE I 
earlier."

We have clarified the definition of the even and odd sectors of the constraint and the choice of $c_P$ in the text.

"The authors conclude that the processes involved in the QDPM 
Hamiltonian are ergodic in each topological sector. This is in marked 
contrast to the QDM on either bipartite or non-bipartite lattice, in 
which we usually have staggered configurations that cannot be reached 
by applying local Hamiltonians repeatedly. Such staggered 
configurations also appear in a recently proposed quantum trimer model 
on a square lattice that exhibits $Z_3$ topological order. I just 
wonder why the authors can exclude such "frozen" configurations from 
the ground states of QDPM. For instance, the state in which all the 
bonds are occupied (all the vertices are of pentamer type) is an 
allowed configuration and does not cost any energy."

We appreciate the referee for pointing our the lack of clarity on this point in the original version of the manuscript. We did not intend to claim that we have proven the the QDPM is ergodic in each topological sector, but simply that our numerics are consistent with the QDPM dynamics connecting disconnected topological sectors in the QDM. We have clarified this in the text by explicitly acknowledging such "frozen" configurations.  Additionally, as mentioned in the manuscript, the "staggered" configuration which are frozen in the QDM do have dynamics in the QDPM (see Fig. 4 in the new version of the manuscript).

@@

"I found several minor typos or grammar errors. I think the authors could revise them and polish the manuscript."

We have corrected several typos based on the referees comments--these corrections are listed below.

"I am a little bit worried about "ergodicity" of the model. For example, a staggered region can hop into a pair pentamer (Table II, v1 case). But, again this pair can not be split into or hop into another configuration including some "flippable" plaquettes in the same topological sector. Therefore, the dynamics still seems to happen only between several (maybe quite a lot in large system sizes, but still does not connect to others) configurations and thus not ergodic. If the authors have concrete answer or evidence, they may add some discussion in details in the manuscript."

We have added a new figure (Fig. 4 in the new manuscript) that illustrates how pentamers can give dynamics to staggered regions, and convert regions with no flippable plaquettes to maximally flippable columnar configurations. We hope this new figure helps illustrate the increased dynamics of the QDPM over the QDM.

@@

--$n_\ell \equiv 1/2+\sigma_\ell^z$ ==> $n_\ell \equiv 1/2(1+\sigma_\ell^z)$

--Definition of even/odd sectors clarified to read:"...where $c_p$ is a constant, and we define the even (odd) sector of the constraint to have $c_P = 2$ ($c_P=1$); the $+$($-$) sign..."

--References to Tab. I and Fig. 1 added

Page 5: Sec. III 2nd Paragraph: ... weaker correlations that the QDM, -> ... weaker correlations than the QDM,

Page 6. Left column: Figure 4 added.

Page 6: Sec III: Rephrased statement about ergodicity: "Our numerical simulations (discussed below in Sec. IV) suggest that the dynamics of $H_{QDPM}$ are ergodic in..." ==> "With the exception isolated ``frozen" configurations (e.g. pentamers occupying all vertices), our numerical simulations (discussed below in Sec. IV) suggest that the dynamics of the $H_{QDPM}$ access a substantial region of configuration space in..."

@@

Page 6: Sec. IV A

--Clarified the discussion of ergodicity as follows:

"Although we cannot rule out the possibility of a variety of isolated ``frozen" configurations (e.g. pentamers occupying all sites) which are not accessed by the dynamics described in \tabref{tab:pentamer_moves}, we find that the increased dynamics of the QDPM connect $U(1)$ topological sectors that are disconnected in the QDM; in particular we note that the QDPM gives dynamics to ``staggered" dimer configurations that are frozen in the QDM (see Fig. 4)."

Page 9: Sec. IV D, 1st paragraph: Removed the following statement: "which appear to provide ergodic dynamics in the monomer-dimer-pentamer configuration space."

### 2017-08-01 qdpm_paper.tex
2nd referee response

Our numerical simulations (discussed below in \secref{sec:numerics}) suggest that the dynamics of $\HQDPM$ are ergodic in each topological sector $\Omega$ that are labeled by the topological loop operators:
\begin{equation}
\Omega = \left(W_X^{Z_3},W_Y^{Z_3} \right).  \notag
\end{equation}
The exact zero energy ground state at the RK point, where $t_i=v_i, i \in \{ 1,2,3\}$, is the RK state, where the superposition is taken to be over all states in the topological sector. Since the $W_{X}^{Z_3}$ and  $W_{Y}^{Z_3}$ have eigenvalues $\{1,e^{i2\pi/3},e^{-i2\pi/3}\}$, the RK point of the QDPM is therefore 9-fold degenerate on the torus.

@@

We explicitly demonstrate that the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$, which is shown in \figref{fig:u1_wind_qdpm}. The different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$, and we find that $\Phi_X$ and $\Phi_Y$ are not  strictly conserved (as they are in the QDM), but are only conserved modulo $3$. The histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors, although we cannot rule out the possibility of isolated ``frozen" configurations which are not accessed by the dynamics described in \tabref{tab:pentamer_moves}; in particular we note that the QDPM gives dynamics to ``staggered" dimer configurations that are frozen in the QDM (see \figref{fig:stag_to_flip}). For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.

@@

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   \notag
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlation length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_parallel.pdf}
    \caption{The dimer-dimer density correlation function $C_d$ between parallel links in the QDPM RK ground state on a $L=64$ lattice, along the direction perpendicular to the link orientation.  The lines represent the best fit to an exponential decay with correlation length $\chi\approx 1$.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To study the confinement of $Z_3$ electrically charged defects in the QDPM, we have Monte Carlo sampled the RK ground state doped with a single pair of monomers. We initialize the system with a monomer on each sub-lattice and supplement the local Monte Carlo updates with updates of the form:
\begin{equation}
\monomermvA  \Longleftrightarrow \monomermvB  \notag
\end{equation}
which appear to provide ergodic dynamics in the monomer-dimer-pentamer configuration space. Notice that monomers remain on the same sub-lattice, which is consistent with their assignment of a $Z_3$ electric charge of $\pm1$ (see \tabref{tab:defects}) and the positive and negative monomers live on different sub-lattices. We then can test for the confinement of monomers by studying the monomer-monomer density correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$, which is plotted in \figref{fig:monomer_correlations}. If the monomers are confined, we expect $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ to decay to zero at large distances; conversely if monomers are deconfined, we expect the monomers to be randomly distributed on the lattice, such that $\langle m_{\bm{0}} m_{\bm{r}} \rangle \simeq L^{-4}$ at large separations. \figref{fig:monomer_correlations} demonstrates that $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ decays to $L^{-4}$ for each system size, indicating a deconfinement of monomers. This is in contrast with the QDM where monomer correlations decay as a power-law. Thus we find that $Z_3$ electric charges are deconfined in the QDPM.
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to $L=64$. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color.}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

With the exception isolated ``frozen" configurations (e.g. pentamers occupying all vertices), our numerical simulations (discussed below in \secref{sec:numerics}) suggest that the dynamics of the $\HQDPM$ access a substantial region of configuration space in each topological sector $\Omega$ that are labeled by the topological loop operators:
\begin{equation}
\Omega = \left(W_X^{Z_3},W_Y^{Z_3} \right).  \notag
\end{equation}
The exact zero energy ground state at the RK point, where $t_i=v_i, i \in \{ 1,2,3\}$, is the RK state, where the superposition is taken to be over all states in the topological sector. Since the $W_{X}^{Z_3}$ and  $W_{Y}^{Z_3}$ have eigenvalues $\{1,e^{i2\pi/3},e^{-i2\pi/3}\}$, the RK point of the QDPM is therefore 9-fold degenerate on the torus.

@@

We explicitly demonstrate that the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$, which is shown in \figref{fig:u1_wind_qdpm}. The different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$, and we find that $\Phi_X$ and $\Phi_Y$ are not  strictly conserved (as they are in the QDM), but are only conserved modulo $3$. The histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. Although we cannot rule out the possibility of a variety of isolated ``frozen" configurations (e.g. pentamers occupying all sites) which are not accessed by the dynamics described in \tabref{tab:pentamer_moves}, we find that the increased dynamics of the QDPM connect $U(1)$ topological sectors that are disconnected in the QDM; in particular we note that the QDPM gives dynamics to ``staggered" dimer configurations that are frozen in the QDM (see \figref{fig:stag_to_flip}). For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.

@@

Additionally we have studied the nature of the decay of the connected dimer-dimer correlation function,
  \begin{equation}
    C_{\mathrm{d}} \left(r\right) \equiv \langle n_{\bm{0}} n_{\bm{r}} \rangle - \langle n_{\bm{0}} \rangle   \langle n_{\bm{r}} \rangle   \notag
\end{equation}
where $r$ is the distance between the links $\bm{0}$ and $\bm{r}$. In \figref{fig:spatial_dmr_cor}, $C_d$ between parallel links is plotted as a function of separation, for $\bm{r}$ perpendicular to the orientation of the dimers for a $L=64$ lattice. We find that $C_d$ decays exponentially with distance having a correlation length  $\chi\approx 1$. Additionally, $C_d$ displays a sub-lattice oscillation that is consistent with the finite-size maximum in $S_d$ at $(0,\pi)$. Thus we find that the RK point of the QDPM is a symmetric dimer liquid with exponentially decaying dimer correlations, in contrast with the power-law decaying correlations of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_parallel.pdf}
    \caption{The dimer-dimer density correlation function $C_d$ between parallel links in the QDPM RK ground state on a $L=64$ lattice, along the direction perpendicular to the link orientation.  The lines represent the best fit to an exponential decay with correlation length $\chi\approx 1$.}
    \label{fig:spatial_dmr_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To study the confinement of $Z_3$ electrically charged defects in the QDPM, we have Monte Carlo sampled the RK ground state doped with a single pair of monomers. We initialize the system with a monomer on each sub-lattice and supplement the local Monte Carlo updates with updates of the form:
\begin{equation}
\monomermvA  \Longleftrightarrow \monomermvB.  \notag
\end{equation}
Notice that monomers remain on the same sub-lattice, which is consistent with their assignment of a $Z_3$ electric charge of $\pm1$ (see \tabref{tab:defects}) and the positive and negative monomers live on different sub-lattices. We then can test for the confinement of monomers by studying the monomer-monomer density correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$, which is plotted in \figref{fig:monomer_correlations}. If the monomers are confined, we expect $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ to decay to zero at large distances; conversely if monomers are deconfined, we expect the monomers to be randomly distributed on the lattice, such that $\langle m_{\bm{0}} m_{\bm{r}} \rangle \simeq L^{-4}$ at large separations. \figref{fig:monomer_correlations} demonstrates that $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ decays to $L^{-4}$ for each system size, indicating a deconfinement of monomers. This is in contrast with the QDM where monomer correlations decay as a power-law. Thus we find that $Z_3$ electric charges are deconfined in the QDPM.
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to $L=64$. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color.}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
